The Talking Heads Experiment

### 2015-04-24 chap8/chap8.tex
adopted corrections by D Schröder and removed subsequent widows & orphans

We did the same for the Talking heads experiments and this was a huge step. Many more people had to get
get involved: for making the code more robust, setting up the teleportation infrastructure, and installing and 
maintaining the physical installations in different countries.  
The mechanisms about language evolution discussed in \partref{part:1} of this book did not significantly change. On the contrary, 
they received solid experimental confirmation. At the same time, we learned a huge amount about the dependencies
between these mechanisms and the environments in which agents use them. We also learned a great deal about running software 
agents in an open infrastructure distributed over the entire globe. Furthermore we learned a lot 
about the interaction between humans and agents ranging from 
enthusiastic participation and enjoyment to nasty attacks by English hackers, set on destroying the experiment.

@@

The Laboratorium\is{Laboratorium exhibition}
exhibition was a major artistic event in Europe during the summer of 1999 organised by 
Bruno Verbergt of an organisation called `Antwerpen Open'.
It was one of the first exhibitions that put so much emphasis on art as a research 
activity and on profound interactions between art and science.
Participants included both scientists and science historians such as Peter Galison, Bruno Latour, Israel Rosenfield
and Isabelle Stengers, as well as architects such as Rem Koolhaas and artists with an affinity for research 
such as Peter Fischli and David Weiss, Gabriel Orozco, Carsten H\"oller, Panamarenko, Lawrence Weiner a.o. 
The exhibition was visited by 8000 people.

@@

The installation in Antwerp was anounced as a ``Laboratory for Cognitive Robotics and Teleportation" \citep{Steels:99b}.
It featured various rooms as shown in the layout in \figref{fig:layout}. The rooms had different functions: 
\begin{enumerate} 
\item The {\itshape central room} which was visible on entering the space contained the experiment itself, i.e. the 
two cameras mounted on tripods, the computers driving them, the screen with geometric figures, and a projection of 
what went on. The activities of the agents were audible through a narration. 
\clearpage
\item A {\itshape reading room} contained background philosophical and scientific 
papers and excerpts from books about language, meaning and their origins. These readings included: 
\item A {\itshape user interface room} contained a workstation where visitors to the exhibition could create their own 
agents, direct them to play games on certain sites, and teach new words to their agents. 
\end{enumerate}

@@

A typical example of the kind of images that the camera's picked up are shown in \figref{fig:images}. 
To play a game, the agent had to find a sufficiently delineated group of objects and each object had \todo{I have framed Figure 8.2 because it contains multiple images}
to exceed a minimal size. Initially quite complex configurations were put on the white board but this made it very difficult for agents to find a coherent group and to develop good concepts for referring to the topic. The right 
image in \figref{fig:images} provides clear examples of up, down, left and right, different colors, and 
also opportunities to use multiple words (such as `red bottom').

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.40\textwidth]{chap8/figures/image-1.jpg}
\includegraphics[width=.40\textwidth]{chap8/figures/image-2.pdf}}
\caption{\label{fig:images}Examples of images picked up by the camera's. The chosen topic was signalled by drawing a bounding box around it. 
During experimentation, it was found that configurations such as in the right image lead to more stable performance.} 
\end{figure}

A typical example of an interaction is shown in \figref{fig:interaction}. 
The speaker (called "rubber", an agent created and named by a user) has selected the blue object at the bottom left 
of the screen. We see the discrimination trees of the agent at that point and the data for each of the objects 
recognised. The coordinates of the respective objects (scaled within the coordinate system of 
the captured image) are (0.0,1.0) for the blue object, (1.0, 0.96) for the red one, and (0.42, 0.0) for the green one. 
The values are displayed as HP (horizontal position), VP (vertical position), H (height), W (width), A (area), 
R (red), G (green), Y (yellow), B (blue), L (lightness). 
The blue object is chosen as the topic and the value on the blue channel has the most discriminative power, although 
there are other possibilities. There are three competing words for naming blue: Xagadude, Nibidesu, and Tetipi. 
None of them have a score higher than 0 and so a random choice of Xagadude is made.

The hearer (called "me") perceives an image which is slightly different from 
the one seen by the speaker. Also the discrimination trees built up so far by this agent are different from the ones used
by the speaker. The hearer does not know the word, so the game fails. After the speaker has then pointed to the 
topic, the hearer conceptualises and guesses that the meaning of this word is 'blue' because that is also for him the most 
discriminating feature of the topic (which has coordinates 0.0, 1.0) compared to the other objects in the context.

@@

The experiment had also a presence on the web, designed by Angus McIntyre. Anyone could log in on the 
Talking Heads website\is{Talking Heads website}, create an agent with 
a given name, and launch it on a tour of the various physical sites. There was also a forum on which users could discuss
the experiment as it was progressing, a hall of fame for the agents that were the best communicators, and 
an overview of the lexicon 
that had formed so far. The website allowed inspection of what was happening at each site (\figref{fig:thwebsite}): 
Users could check which agents were waiting there to play a game, and what the current 
game was about. It displayed also statistics about the 
communicative success and agent activity at each site. The site is no longer operational due to 
changes to the underlying software but some remnants can be visited here: https://ai.vub.ac.be/talking-heads/

@@

	A
\end{mail}
\begin{mail}
From: Angus McIntyre <angus@csl.sony.fr>\\
Subject: Re: Database cleared\\

@@

I'm off home now. If you need me to come back and kick 
the server, call me on:
	+33-1-42-78-73-88
Bye,
	A
\end{mail}

@@

By the way, there's a bug in the 'headnet' software with respect to 
routes; there's a finite limit on the field length for the 'route' 
field. If you make a route that's too long, it gets truncated, so 
you can have an agent whose route goes:
	... paris 10 bruss
The agent will complete its games on 'paris' and then go into 
stasis waiting for a server called 'bruss' to pick it up. A patch 
for this would probably be nice to have at some point.

@@

3. Masking. Eliminating zero channels should be done on the source not on
value (= value after scaling, so it is often zero). corrected in: 
prototype.lsp (in the functions folder)

After theses patches, sites can be running again.

Fred and Luc.
\end{mail}
\clearpage
\begin{mail}
--============\_-1278188520==\_d============\\
Date: Tue, 29 Jun 1999 23:25:45 +0200 (MET DST)\\
Subject: There is hope for robokind \\

@@

By mid July, both the basic infrastructure, the teleportation mechanisms and the semiotic dynamics were running 
smoothly and so the experiment could operate without constant care.

@@

\figref{fig:woggle} shows the semiotic dynamics related to an expression of the meaning for the 
concept `left', i.e. a horizontal position 
between 0.0 and 0.25 (after scaling). The word ``wogglesplat" becomes dominant
although there is a very strong competition in the beginning. 
We see that users try to give other words to the same meaning, such as "gauche" or "links" (both words expressing left 
in French and German or Dutch respectively). We also see that other words such as "red" or "yellow" get associated with this 
meaning, because the hearer may guess the wrong meaning in learning the word for left.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/wogglesplat.pdf}}
\caption{\label{fig:woggle}Meaning-Form diagram: Different words for expressing the meaning 'left', i.e. 
horizontal position is less than 0.25 (scaled). 
New words come up all the time but there is a clear winner-take-all effect with "wogglesplat" winning. 
}
\end{figure}
\enlargethispage{1\baselineskip}
\figref{fig:droite} shows another example of the semiotic dynamics in the experiment, this time looking at all the 
meanings for a particular word, namely `droite' (meaning `to the right' in French). This word clearly has been introduced
by a human user and the dominating meaning progressively becomes a region (between 0.75 and 1.0) on the horizontal position, 
as could be expected. On the way we see some confusion, Specifically there must be objects that appear both on the right 
and up (vertical position between 0.5 and 1.0).

@@

Numerous talks were given and various papers published in scientific fora. There were also various talks within 
the art context.\footnote{Examples are a `gallery talk' as Salon 3 at the Elephant and Castle Centre
in London on 2 December 1998, organised by Hans Ulrich Obrist and Molly Nesbit, and a presentation together with 
Hans-Ulrich Obrist and Rem Koolhaas in Antwerp to launch the Laboratorium book on 3 October 2001.}
The experiment received wide coverage in 
the media thanks to its public exposure as part of a major exhibition. The Süddeutsche Zeitung called it 
``Angels with Internet wings''. All this lead to invitations to show the experiment
also in other venues, as discussed in the next chapter. 
Regretably, the challenge just to keep the experiment in the air with the available human resources
and the pressure for going on with new experiments prevented us from doing more adequate data gathering 
and analysis. Nevertheless some analyses appeared, particularly as carried out by Frederic Kaplan (\citealt{Kaplan:2001}).

The field of complex systems science was at that time still in its infance and 
adequate tools for analysing language as a complex adaptive system were in the early stages of development. 
Nevertheles, various publications were

The experiment ended with the following ``tongue-in-cheek'' post by Angus McIntyre on the Forum:

\begin{mail}
{\bfshape 1999-10-14 15:08:29 Angus McIntyre}\\
{\itshape Bad news and good news}\\

>>>>

We did the same for the Talking Heads experiments and this was a huge step. Many more people had to get
get involved: for making the code more robust, setting up the teleportation infrastructure, and installing and 
maintaining the physical installations in different countries.  
The mechanisms about language evolution discussed in \partref{part:1} of this book did not significantly change. On the contrary, 
they received solid experimental confirmation. At the same time, we learned a huge amount about the dependencies
between these mechanisms and the environments in which agents use them. We also learned a great deal about running software 
agents in an open infrastructure distributed over the entire globe. Furthermore we learned a lot 
about the interaction between humans and agents ranging from 
enthusiastic participation and enjoyment to nasty attacks by English hackers, set on destroying the experiment.

@@

The Laboratorium\is{Laboratorium exhibition}
exhibition was a major artistic event in Europe during the summer of 1999 organised by 
Bruno Verbergt of an organisation called `Antwerpen Open'.
It was one of the first exhibitions that put so much emphasis on art as a research 
activity and on profound interactions between art and science.
Participants included both scientists and science historians such as Peter Galison, Bruno Latour, Israel Rosenfield
and Isabelle Stengers, as well as architects such as Rem Koolhaas and artists with an affinity for research 
such as Peter Fischli and David Weiss, Gabriel Orozco, Carsten H\"oller, Panamarenko, Lawrence Weiner and others. 
The exhibition was visited by 8000 people.

@@

The installation in Antwerp was anounced as a ``Laboratory for Cognitive Robotics and Teleportation" \citep{Steels:99b}.
It featured various rooms as shown in the layout in \figref{fig:layout}. The rooms had different functions: 
\begin{enumerate} 
\item The {\itshape central room} which was visible on entering the space contained the experiment itself, i.e. the 
two cameras mounted on tripods, the computers driving them, the screen with geometric figures, and a projection of 
what went on. The activities of the agents were audible through a narration. 
\clearpage
\item A {\itshape reading room} contained background philosophical and scientific 
papers and excerpts from books about language, meaning and their origins. These readings included: \todo{The readings are referred to, but they do not follow after the colon}
\item A {\itshape user interface room} contained a workstation where visitors to the exhibition could create their own 
agents, direct them to play games on certain sites, and teach new words to their agents. 
\end{enumerate}

@@

A typical example of the kind of images that the cameras picked up are shown in \figref{fig:images}. 
To play a game, the agent had to find a sufficiently delineated group of objects and each object had \todo{I have framed Figure 8.2 because it contains multiple images}
to exceed a minimal size. Initially quite complex configurations were put on the white board but this made it very difficult for agents to find a coherent group and to develop good concepts for referring to the topic. The right 
image in \figref{fig:images} provides clear examples of up, down, left and right, different colors, and 
also opportunities to use multiple words (such as `red bottom').

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.40\textwidth]{chap8/figures/image-1.jpg}
\includegraphics[width=.40\textwidth]{chap8/figures/image-2.pdf}}
\caption{\label{fig:images}Examples of images picked up by the cameras. The chosen topic was signalled by drawing a bounding box around it. 
During experimentation, it was found that configurations such as in the right image lead to more stable performance.} 
\end{figure}

A typical example of an interaction is shown in \figref{fig:interaction}. 
The speaker (called ``rubber", an agent created and named by a user) has selected the blue object at the bottom left 
of the screen. We see the discrimination trees of the agent at that point and the data for each of the objects 
recognised. The coordinates of the respective objects (scaled within the coordinate system of 
the captured image) are (0.0,1.0) for the blue object, (1.0, 0.96) for the red one, and (0.42, 0.0) for the green one. 
The values are displayed as \textsc{hp} (horizontal position), \textsc{vp} (vertical position), \textsc{h} (height), \textsc{w} (width), \textsc{a} (area), 
\textsc{r} (red), \textsc{g} (green), \textsc{y} (yellow), \textsc{b} (blue), \textsc{l} (lightness). 
The blue object is chosen as the topic and the value on the blue channel has the most discriminative power, although 
there are other possibilities. There are three competing words for naming blue: `Xagadude', `Nibidesu', and `Tetipi'. 
None of them have a score higher than 0 and so a random choice of `Xagadude' is made.

The hearer (called ``me") perceives an image which is slightly different from 
the one seen by the speaker. Also the discrimination trees built up so far by this agent are different from the ones used
by the speaker. The hearer does not know the word, so the game fails. After the speaker has then pointed to the 
topic, the hearer conceptualises and guesses that the meaning of this word is `blue' because that is also for him the most 
discriminating feature of the topic (which has coordinates 0.0, 1.0) compared to the other objects in the context.

@@

The experiment had also a presence on the web, designed by Angus McIntyre. Anyone could log in on the 
Talking Heads website\is{Talking Heads website}, create an agent with 
a given name, and launch it on a tour of the various physical sites. There was also a forum on which users could discuss
the experiment as it was progressing, a hall of fame for the agents that were the best communicators, and 
an overview of the lexicon 
that had formed so far. The website allowed inspection of what was happening at each site (\figref{fig:thwebsite}): 
Users could check which agents were waiting there to play a game, and what the current 
game was about. It also displayed statistics about the 
communicative success and agent activity at each site. The site is no longer operational due to 
changes to the underlying software but some remnants can be visited here: \url{https://ai.vub.ac.be/talking-heads/}

@@

	A \enlargethispage{1\baselineskip}
\end{mail}
\begin{mail}
From: Angus McIntyre <angus@csl.sony.fr>\\
Subject: Re: Database cleared\\

@@

I'm off home now. If you need me to come back and kick 
the server, call me on:
	+33-1-42-78-73-88 \todo{If this was a private number, we would not show it here.}
Bye,
	A
\end{mail}

@@

By the way, there's a bug in the 'headnet' software with respect to 
routes; there's a finite limit on the field length for the 'route' 
field. If you make a route that's too long, it gets truncated, so 
you can have an agent whose route goes:\\
	... paris 10 bruss\clearpage
The agent will complete its games on 'paris' and then go into 
stasis waiting for a server called 'bruss' to pick it up. A patch 
for this would probably be nice to have at some point.

@@

3. Masking. Eliminating zero channels should be done on the source not on
value (= value after scaling, so it is often zero). corrected in: 
prototype.lsp (in the functions folder)
\clearpage
After theses patches, sites can be running again.

Fred and Luc.
\end{mail}
\begin{mail}
--============\_-1278188520==\_d============\\
Date: Tue, 29 Jun 1999 23:25:45 +0200 (MET DST)\\
Subject: There is hope for robokind \\

@@

By mid July,\todo{removed `both' here} the basic infrastructure, the teleportation mechanisms and the semiotic dynamics were running 
smoothly and so the experiment could operate without constant care.

@@

\figref{fig:woggle} shows the semiotic dynamics related to an expression of the meaning for the 
concept `left', i.e. a horizontal position 
between 0.0 and 0.25 (after scaling). The word ``wogglesplat" becomes dominant
although there is a very strong competition in the beginning. 
We see that users try to give other words to the same meaning, such as "gauche" or "links" (both words expressing left 
in French and German or Dutch respectively). We also see that other words such as "red" or "yellow" get associated with this \enlargethispage{1\baselineskip}
meaning, because the hearer may guess the wrong meaning in learning the word for left.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/wogglesplat.pdf}}
\caption{\label{fig:woggle}Meaning-Form diagram: Different words for expressing the meaning 'left', i.e. 
horizontal position is less than 0.25 (scaled). 
New words come up all the time but there is a clear winner-take-all effect with "wogglesplat" winning. 
}
\end{figure}

\figref{fig:droite} shows another example of the semiotic dynamics in the experiment, this time looking at all the 
meanings for a particular word, namely `droite' (meaning `to the right' in French). This word clearly has been introduced
by a human user and the dominating meaning progressively becomes a region (between 0.75 and 1.0) on the horizontal position, 
as could be expected. On the way we see some confusion, Specifically there must be objects that appear both on the right 
and up (vertical position between 0.5 and 1.0).

@@

Numerous talks were given and various papers published in scientific fora. There were also various talks within 
the art context.\footnote{Examples are a `gallery talk' as Salon 3 at the Elephant and Castle Centre
in London on 2 December 1998, organised by Hans Ulrich Obrist and Molly Nesbit, and a presentation together with 
Hans-Ulrich Obrist and Rem Koolhaas in Antwerp to launch the Laboratorium book on 3 October 2001.}
The experiment received wide coverage in 
the media thanks to its public exposure as part of a major exhibition. The Süddeutsche Zeitung called it 
``Angels with Internet wings''. All this lead to invitations to show the experiment
also in other venues, as discussed in the next chapter. 
Regrettably, the challenge just to keep the experiment in the air with the available human resources
and the pressure for going on with new experiments prevented us from doing more adequate data gathering 
and analysis. Nevertheless some analyses appeared, particularly as carried out by Frederic Kaplan (\citealt{Kaplan:2001}).

The field of complex systems science was at that time still in its infancy and 
adequate tools for analysing language as a complex adaptive system were in the early stages of development. 
Nevertheles, various publications were \todo{Here the rest of the sentence seems to have disappeared.}

The experiment ended with the following ``tongue-in-cheek'' post by Angus McIntyre on the Forum:
\clearpage
\begin{mail}
{\bfshape 1999-10-14 15:08:29 Angus McIntyre}\\
{\itshape Bad news and good news}\\
